Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2021-11-06 analysis/paper_draft2.Rmd
revised to asa format; figure widths; revised references

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - Sachin S. Pandya
#    orcid_id: 0000-0001-7387-1307
  - Xiaomeng Li
  - Timothy E. Moore
#    orcid_id: 0000-0002-9576-0517
#output: distill::distill_article
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_width: 5
    fig_height: 5
header-includes: 
  - \usepackage{amsmath}
fontsize: 12pt
bibliography: references.bib
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with actual historical strike data from a convenience sample of cases from one federal district court.

@@

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or membership in another cognizable class. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not [@LaFave2020, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class membership was a "factor" in that party's use of strikes [Wash. Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases. (_Flowers v. Mississippi_ -@flowers2019, p. 2243; Wash. Rule 37(g)(v); Calif. Code of Civil Procedure $\S$ 231.7(d)(3)(G)). Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Ibrahim2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter.  Finally, we use that posterior distribution as the prior for the bias parameter in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero; when $b_{curr}$ equals to zero, the credible interval includes zero; and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{current}$ towards zero, we focus not on how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias.

@@

Figures 1-3 depicts the coverage rates of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively. Here, $b_{hist}$ denotes the bias parameter for the generated historical strike data and $b_{curr}$ denotes the bias parameter for the current data. The values on the right side represent the size of the historical data. The values on the top of the plots represent the power prior weight parameter $a$. For any given combination, a darker color shade denotes a higher coverage rate, and vice versa.

@@

load_path <- here::here("analysis","figures","C80plot.png")
ggsave(load_path, plot = C80plot)
load_path <- here::here("analysis","figures","C90plot.png")
ggsave(load_path, plot = C90plot)
load_path <- here::here("analysis","figures","C95plot.png")
ggsave(load_path, plot = C95plot)

@@

```{r fig1, fig.cap = "Coverage Rates - 80% Credible Interval", fig.align = "center"}

@@

```{r fig2, fig.cap = "Coverage Rates - 90% Credible Interval", fig.align = "center"}

@@

```{r fig3, fig.cap = "Coverage Rates - 95% Credible Interval", fig.align = "center"}

@@

Figure 1 depicts the coverage rate with an 80\% credible interval. As the first column ($a = 0.1$) depicts, when we place low weight on the historical strike data, we mostly rely on current information. When $b_{current}$ is 3, 2 or -2, -3, the coverage rate is high, very close to 1. However, when $b_{curr}$ is 1, the coverage rate drops below 0.8. If the bias is not strong, our model detects it with a lower accuracy.

As we place more weight on the historical strike data (as $a$ increases), when the historical strike data is consistent with current information, the accuracy of detecting the bias is also close to 1 when the bias parameter is 1. When the bias parameter is 0 (no bias), however, the coverage rate is also around 80\%. Thus, if we use the 80\% credible interval, we have about a 20\% chance of erroneously detecting non-zero bias.

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 2), when $b = 2$, the coverage rate is lower (the color is lighter) than with a 80\% credible interval; when $b=0$, the color is darker (higher coverage rate). The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

Figure 3 depicts coverage rates when using a 95\% credible interval. Here, given $b_{curr} = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to infer bias when there is none. When $b_{curr} = 3$, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b_{curr} = \{2, -2\}$), the coverage rate is only high when we have consistent historical information.

@@

```{r fig4}

#| fig.cap="Simulation results for simultaneous strikes example"

@@

Figure 4 depicts the results. The columns are values of bias used to simulate the data. Blue points denote estimated posterior mean for the bias parameter. Blue lines denote the 95% credible interval. Red lines denote true value of bias parameters in the simulations.

We find the estimated bias parameter of the 50 trials with different strike orders are close to each other for all of the three scenarios. None of their credible intervals include zero. We therefore infer that the order of strikes does not influence the estimate of the bias parameter. This confirms that the simultaneous-strikes process can be modeled as a special case of our model of an alternating-strikes process.

@@

```{r inpts}
df <- data.frame(round = c(1:10),
                 num_cog = c(5,5,4,3,2,2,2,2,2,2),
                 total = c(21,20,19,18,17,16,15,14,13,12),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))
knitr::kable(df, caption ="Hypothetical Strike Tally")

@@

```{r figapp1, out.width='50%', fig.align='center'}

#| fig.cap = "Density Plots without Historical Strike Data"

@@

Finally, if we assign minimal weight to the historical strike data, the density plot updates accordingly (Figure 8).

@@

At the same time, the Bayesian approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognitive-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike. Moreover, suppose an attorney of interest exercised strikes in two courts, each with different strike procedures. In some cases, one court's strike procedure can be modeled as a special case of a more general model that applied to other court's procedure (as we showed above). If not, that attorney's historical strike data in one court could not used to estimate the posterior of that attorney's bias parameter, given a current trial in the other court.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output: 
  rticles::asa_article:
    citation_package: "default"
    keep_tex: TRUE
    #fig_width: 5
    #fig_height: 5
  #output:
#  pdf_document:
#    latex_engine: xelatex
#    number_sections: true
    #fig_width: 4
    #fig_height: 4
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with actual historical strike data from a convenience sample of cases from one federal district court.

@@

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or membership in another cognizable class. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2020, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class membership was a "factor" in that party's use of strikes [Wash. Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. Rule 37(g)(v); Calif. Code of Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter.  Finally, we use that posterior distribution as the prior for the bias parameter in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero; when $b_{curr}$ equals to zero, the credible interval includes zero; and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus not on how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias.

@@

Figures 1-3 depicts the coverage rates of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively, for the combinations of $b_{hist}$, $b_{curr}$, the power prior weight parameter $a$, and different sizes of the historical data ($h = {1,2,3}$). For any given combination, a darker color shade denotes a higher coverage rate, and vice versa.

@@

load_path <- here::here("analysis","figures","C80plot.png")
ggsave(load_path, plot = C80plot, height = 4, width = 7)
load_path <- here::here("analysis","figures","C90plot.png")
ggsave(load_path, plot = C90plot, height = 4, width = 7)
load_path <- here::here("analysis","figures","C95plot.png")
ggsave(load_path, plot = C95plot, height = 4, width = 7)

@@

```{r fig1, out.width = "95%", fig.cap = "Coverage Rate of bias parameter based on 80 pct credible interval."}

@@

Figure 1 depicts the coverage rate with an 80\% credible interval. When we place low weight on the historical strike data ($a = 0.1$), we mostly rely on current strike information. When $b_{curr}$ is 3, 2, -2, or -3, the coverage rate is high, very close to 1. However, when $b_{curr} = 1$, the coverage rate drops below 0.8. If the bias is not strong, our model detects it with a lower accuracy.

As we place more weight on the historical strike data (as $a$ increases), when the historical strike data is consistent with current information, the accuracy of detecting the bias is also close to 1 when $b_{curr} = 1$. When $b_{curr} = 0$ (no bias), however, the coverage rate is also around 80\%. Thus, if we use the 80\% credible interval, we have about a 20\% chance of erroneously detecting non-zero bias.

```{r fig2, out.width = "95%", fig.cap = "Coverage Rates - 90 pct. Credible Interval"}

@@

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 2), when $b_{curr} = 2$, the coverage rate is lower than with an 80\% credible interval; when $b_{curr}=0$, the coverage rate is higher. The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

```{r fig3, out.width = "95%", fig.cap = "Coverage Rates - 95 pct. Credible Interval"}

@@

Figure 3 depicts coverage rates when using a 95\% credible interval. Here, given $b_{curr} = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to infer bias when there is none. When $b_{curr} = 3$, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b_{curr} = \{2, -2\}$), the coverage rate is only high if we have consistent historical information.

@@

```{r fig4, out.width = "75%", fig.cap="Simulation results for simultaneous strikes example. Columns correspond to values of bias used to simulate data. Points and blue lines denote estimated posterior mean and 95% credible interval for bias parameter. Red line: true value of the bias parameters used in the 50 simulation s."}

@@

Figure 4 depicts the results. We find the estimated bias parameter of the 50 trials with different strike orders are close to each other for all of the three scenarios. None of their credible intervals include zero. We therefore infer that the order of strikes does not influence the estimate of the bias parameter. This confirms that the simultaneous-strikes process can be modeled as a special case of our model of an alternating-strikes process.

@@

```{r inpts}
df <- data.frame(round = c(1:10),
                 num_cog = c(5,5,4,3,2,2,2,2,2,2),
                 total = c(21,20,19,18,17,16,15,14,13,12),
                 cog = c(0,1,1,0,0,1,1,0,0,0),
                 party = rep(c("PP","PD"),5))

knitr::kable(df, caption ="Hypothetical Strike Tally", format = "latex")

@@

```{r figapp1, out.width='50%', fig.cap = "Density Plots without Historical Strike Data"}

@@

Finally, if we assign minimal weight to the historical strike data ($a = 0.1$), the density plot updates accordingly (Figure 8).

@@

At the same time, the Bayesian approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognitive-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike. Moreover, suppose an attorney of interest exercised strikes in two courts, each with different strike procedures. In some cases, one court's strike procedure can be modeled as a special case of a more general model that applied to another court's procedure (as we showed above). If not, that attorney's historical strike data in one court could not used to estimate the posterior of that attorney's bias parameter, given a current trial in the other court.
